---
title: ""
author: ""
date: ""
header-includes:
  - \usepackage{dcolumn}
  - \usepackage{setspace}
  - \usepackage{float}
  - \usepackage{placeins}
output: 
  pdf_document:
    latex_engine: xelatex
---




\begin{center}
{\LARGE \textbf{The Effects of the First European Carbon Taxes on Carbon Emissions}}\\[1.5em]

{\large Evan Toth}\\[0.5em]
{\large Lawrence University}\\[0.5em]
{\large Faculty Advisors: Andrew Sage, Abhishek Chakraborty, Jonathan Lhost}\\[1.5em]

{\large March 18, 2026}
\end{center}

\vspace{2em}




```{r setup, include=FALSE}
# knitting settings
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning=FALSE)
```

```{r, warning=FALSE, include=FALSE}
# loading in the packages
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
library(broom)
library(knitr)
library(fixest)
library(forcats)
library(kableExtra)
library(stargazer)
library(fixest)
library(modelsummary)
```


```{r,include=FALSE}
# first, beginning with temperature data
temp_data <- read_csv("data/monthly-average-surface-temperatures-by-year.csv")

temp_clean <- temp_data %>%
  filter(!is.na(Code))
```


```{r, include=FALSE}
# pivoting and cleaning the data for the desired structure
temp_long <- temp_clean %>%
  pivot_longer(
    cols = matches("^[0-9]{4}$"),
    names_to = "year",
    values_to = "temp_c"
  ) %>%
  mutate(
    month = as.integer(Year),
    year = as.integer(year)
  ) %>%
  select(Entity, Code, year, month, temp_c)

```

```{r, include=FALSE}
# calculating average yearly temperature
temp_annual <- temp_long %>%
  group_by(Entity, Code, year) %>%
  summarize(
    annual_temp_c = mean(temp_c, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, include=FALSE}
# removing any 2025 data (not all values were available for all locations)
temp_annual <- temp_annual %>%
  filter(year <= 2024)
```



```{r, include=FALSE}
# democracy data
DI_data <- read_csv("data/EIU_DI.csv")
```

```{r, include=FALSE}
# filtering out continents
demo_clean <- DI_data %>%
  filter(str_detect(REF_AREA, "^[A-Z]{3}$"))
```

```{r, include=FALSE}
# selecting the needed variables to join
demo_clean <- demo_clean %>%
  select(
    Code = REF_AREA,
    Country = REF_AREA_LABEL,
    Year = TIME_PERIOD,
    DemocracyIndex = OBS_VALUE
  )
```


```{r, include=FALSE}
# joining democracy index data with temperature data

demotemp_data <- temp_annual %>%
  left_join(demo_clean, by = c("Code" = "Code", "year" = "Year"))
```

```{r, include=FALSE}
# removing the duplicate country generated by the join
demotemp_data <- demotemp_data %>% select(-Country)
```


```{r, include=FALSE}
# now GDP data
gdp_data <- read_csv("data/GDPData.csv")
```

```{r, include=FALSE}
# cleaning and pivoting the data into the desired structure for joining
gdp_long_clean <- gdp_data %>% 
  filter(`Series Name` == "GDP per capita (constant 2015 US$)") %>% 
  pivot_longer(
    cols = matches("^[0-9]{4}"),
    names_to = "year",
    values_to = "GDP_per_capita"
  ) %>% 
  mutate(
    year = as.integer(sub(" .*", "", year)),
    GDP_per_capita = as.numeric(GDP_per_capita)
  ) %>% 
  select(`Country Code`, year, GDP_per_capita)
```


```{r, include=FALSE}
# joining GDP data with the other 2 controls
control_data <- demotemp_data %>% 
  left_join(
    gdp_long_clean,
    by = c("Code" = "Country Code", "year" = "year")
  )
```



```{r, include=FALSE}
# manufacturing data
manufacturing_data <- read_csv("data/Manufacturingdata.csv")
```


```{r, include=FALSE}
# manufacturing (% of GDP) data (already read in as manufacturing_data)
mfg_long_clean <- manufacturing_data %>%
  pivot_longer(
    cols = matches("^\\d{4} \\[YR\\d{4}\\]$"),
    names_to = "year",
    values_to = "mfg_va_pct_gdp"
  ) %>%
  mutate(
    # extract the 4-digit year from strings like "1990 [YR1990]"
    year = as.integer(str_extract(year, "^\\d{4}")),
    
    # convert ".." to NA, then numeric
    mfg_va_pct_gdp = na_if(mfg_va_pct_gdp, ".."),
    mfg_va_pct_gdp = as.numeric(mfg_va_pct_gdp)
  ) %>%
  filter(year >= 1985, year <= 2024) %>%
  select(`Country Code`, year, mfg_va_pct_gdp)

```


```{r, include=FALSE}
control_data <- control_data %>%
  left_join(
    mfg_long_clean,
    by = c("Code" = "Country Code", "year" = "year")
  )
```

```{r, include=FALSE}
# trade openness data
trade <- read_csv("data/TradeOpennessData.csv")
```


```{r, include=FALSE}
# cleaning and pivoting trade data for joining
trade_long_clean <- trade %>%
  pivot_longer(
    cols = matches("^\\d{4} \\[YR\\d{4}\\]$"),
    names_to = "year",
    values_to = "tradepercent"
  ) %>%
  mutate(
    year = as.integer(str_extract(year, "^\\d{4}")),
    tradepercent = na_if(tradepercent, ".."),
    tradepercent = as.numeric(tradepercent)
  ) %>%
  filter(year >= 1985, year <= 2024) %>%
  select(`Country Code`, year, tradepercent)

```


```{r, include=FALSE}
# joining trade data in with other controls
control_data <- control_data %>%
  left_join(
    trade_long_clean,
    by = c("Code" = "Country Code", "year" = "year")
  )
```


```{r, include=FALSE}
# Net Import and Export Data
importsexports <- read_csv("data/ImportsExports(1).csv")
```


```{r, include=FALSE}
# pivoting import and export data
year_cols <- names(importsexports)[
  stringr::str_detect(names(importsexports), "^\\d{4} \\[YR\\d{4}\\]$")
]

impexp_long <- importsexports %>%
  mutate(across(all_of(year_cols), as.character)) %>%   # <- key fix
  pivot_longer(
    cols = all_of(year_cols),
    names_to = "year",
    values_to = "value"
  ) %>%
  mutate(
    year  = as.integer(str_extract(year, "^\\d{4}")),
    value = na_if(value, ".."),
    value = as.numeric(value)
  ) %>%
  select(`Country Code`, `Series Code`, year, value)
```

```{r, include=FALSE}
# widening for joining
impexp_wide <- impexp_long %>%
  filter(`Series Code` %in% c("NE.IMP.GNFS.ZS", "NE.EXP.GNFS.ZS")) %>%
  pivot_wider(
    names_from = `Series Code`,
    values_from = value
  ) %>%
  rename(
    imports_gdp = `NE.IMP.GNFS.ZS`,
    exports_gdp = `NE.EXP.GNFS.ZS`
  )
```


```{r, include=FALSE}
# joining export and import data
control_data <- control_data %>%
  left_join(impexp_wide, by = c("Code" = "Country Code", "year" = "year"))
```


```{r, include=FALSE}
# Fossil Fuel Data
fossil_raw <- read_csv("data/FossilFuelData.csv")

# helper to avoid NaN means later
safe_mean <- function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)

fossil_long_clean <- fossil_raw %>%
  pivot_longer(
    cols = matches("^\\d{4} \\[YR\\d{4}\\]$"),
    names_to = "year",
    values_to = "fossil_share"
  ) %>%
  mutate(
    year = as.integer(str_extract(year, "^\\d{4}")),
    fossil_share = na_if(fossil_share, ".."),
    fossil_share = as.numeric(fossil_share)
  ) %>%
  filter(year >= 1985, year <= 2024) %>%
  select(`Country Code`, year, fossil_share)

control_data <- control_data %>%
  left_join(
    fossil_long_clean,
    by = c("Code" = "Country Code", "year" = "year")
  )

```



```{r, include=FALSE}
# carbon emissions data
co2_data <- read_csv("data/co-emissions-per-capita.csv")
```

```{r, include=FALSE}
# cleaning data for joining

co2_clean <- co2_data %>%
  # 1. Remove continents (they have missing country codes)
  filter(!is.na(Code)) %>%
  
  # 2. Keep years past 1950
  filter(Year >= 1950) %>%
  
  # 3. Rename variables
  rename(
    co2_per_capita = `Annual CO₂ emissions (per capita)`,
    year = Year
  ) %>%
  
  # 4. Keep only what is needed
  select(Code, year, co2_per_capita)
```




```{r, include=FALSE}
# joining carbon emission data with control data
data_without_policy <- control_data %>%
  left_join(co2_clean, by = c("Code", "year"))
```


```{r, include=FALSE}
# carbon tax data
tax_data <- read_csv("data/carbontaxdata.csv")
```

```{r, include=FALSE}
# cleaning and preparing tax data for joining
tax_clean <- tax_data %>%
  mutate(
    carbon_tax = case_when(
      `Covered by tax instrument in at least one sector` == "No carbon tax" ~ 0,
      TRUE ~ 1
    )
  ) %>%
  select(Code, Year, carbon_tax)
```

```{r, include=FALSE}
# renaming year variable for joining
tax_clean <- tax_clean %>%
  rename(year = Year)
```

```{r, include=FALSE}
# joining tax with other data
with_tax_data <- data_without_policy %>%
  left_join(tax_clean, by = c("Code", "year"))
```


```{r, include=FALSE}
# ETS data now
ETS_Data <- read_csv("data/ETSData.csv")
```

```{r, include=FALSE}
# cleaning the ETS data, prepping for join
ets_clean <- ETS_Data %>%
  mutate(
    ets = case_when(
      `Covered by emissions trading system in at least one sector` == "No ETS" ~ 0,
      TRUE ~ 1
    )
  ) %>%
  select(Code, Year, ets)
```

```{r, include=FALSE}
# renaming year variable for join
ets_clean <- ets_clean %>%
  rename(year = Year)
```

```{r, include=FALSE}
# creating final prepped dataset by joining ETS data in
final_data <- with_tax_data %>%
  left_join(ets_clean, by = c("Code", "year")) %>%
  filter(year >= 1985) # filtering out any year before 1985
```


```{r, include = FALSE}
# replacing NA values for the carbon tax and ets indicators with 0 
final_data <- final_data %>%
  mutate(
    carbon_tax = if_else(is.na(carbon_tax), 0L, carbon_tax),
    ets        = if_else(is.na(ets), 0L, ets)
  )
```

```{r, include = FALSE}
# calculating a first tax year for each country that has a carbon tax in preparation for the event study
tax_years <- final_data %>%
  group_by(Entity, Code) %>%
  summarise(
    first_tax_year = if (any(carbon_tax == 1, na.rm = TRUE)) {
      min(year[carbon_tax == 1], na.rm = TRUE)
    } else {
      NA_integer_
    },
    .groups = "drop"
  ) %>%
  mutate(
    early_adopter_1990_2004 = as.integer(
      !is.na(first_tax_year) &
      first_tax_year >= 1990 &
      first_tax_year <= 2004
    )
  )
```




```{r, include = FALSE}
# keeping the 29 chosen countries for the study
keep_codes <- c(
  "AUT","BEL","BGR","HRV","CYP","CZE","DNK","EST","FIN","FRA","DEU",
  "GRC","HUN","IRL","ITA","LVA","LTU","NLD","POL","PRT","ROU","SVK",
  "SVN","ESP","SWE","GBR","NOR","CHE","ISL"
)

analysis_data_europe <- final_data %>%
  filter(Code %in% keep_codes) %>%
  select(-DemocracyIndex)
```


```{r}
analysis_data_europe <- analysis_data_europe %>%
  mutate(
    fossil_share = case_when(
      fossil_share < 0 ~ NA_real_,
      fossil_share > 100 ~ NA_real_,
      fossil_share == 0 ~ NA_real_,   # <-- your idea, generalized
      TRUE ~ fossil_share
    )
  )
```




```{r, include = FALSE}
# creating region variables for each country
region_map <- analysis_data_europe %>%
  distinct(Entity, Code) %>%
  mutate(
    region = case_when(
      Entity %in% c("Norway", "Sweden", "Denmark", "Finland", "Iceland") ~ "Nordic",
      Entity %in% c("Germany", "France", "Austria", "Belgium",
                    "Netherlands", "Luxembourg") ~ "Western Europe",
      Entity %in% c("Italy", "Spain", "Portugal", "Greece") ~ "Southern Europe",
      Entity %in% c("Poland", "Czechia", "Slovakia", "Hungary",
                    "Slovenia", "Croatia", "Estonia", "Latvia", "Lithuania") ~ "Central & Eastern Europe",
      Entity %in% c("United Kingdom", "Switzerland") ~ "Non-EU Europe",
      TRUE ~ "Other"
    )
  )
```



```{r, include = FALSE}
# making a function to calculate average levels over 5 year gaps for countries
# in preparation for calculating propensity scores
make_baseline_fixed_window <- function(df, start_year, end_year, window_label) {

  base_means <- df %>%
    filter(year >= start_year, year <= end_year) %>%
    group_by(Entity, Code) %>%
    summarise(
      baseline_trade   = mean(tradepercent, na.rm = TRUE),
      baseline_exports = mean(exports_gdp, na.rm = TRUE),
      baseline_imports = mean(imports_gdp, na.rm = TRUE),
      baseline_mfg     = mean(mfg_va_pct_gdp, na.rm = TRUE),
      baseline_gdp     = mean(GDP_per_capita, na.rm = TRUE),
      baseline_co2     = mean(co2_per_capita, na.rm = TRUE),
      baseline_temp     = mean(annual_temp_c, na.rm = TRUE),

      trade_obs   = sum(!is.na(tradepercent)),
      exports_obs = sum(!is.na(exports_gdp)),
      imports_obs = sum(!is.na(imports_gdp)),
      mfg_obs     = sum(!is.na(mfg_va_pct_gdp)),
      gdp_obs     = sum(!is.na(GDP_per_capita)),
      co2_obs     = sum(!is.na(co2_per_capita)),
      temp_obs     = sum(!is.na(annual_temp_c)),

      .groups = "drop"
    )

  co2_trend <- df %>%
    filter(year >= start_year, year <= end_year) %>%
    filter(!is.na(co2_per_capita)) %>%
    group_by(Entity, Code) %>%
    summarise(
      co2_trend_baseline = if (n() >= 2) coef(lm(co2_per_capita ~ year))[["year"]] else NA_real_,
      .groups = "drop"
    )

  base_means %>%
    left_join(co2_trend, by = c("Entity", "Code")) %>%
    mutate(baseline_window = window_label)
}
```


```{r, include = FALSE}
# creating a dataset with 5 year averages over 1985-89, and 1990-94
baseline_85_89 <- make_baseline_fixed_window(
  df = analysis_data_europe,
  start_year = 1985,
  end_year = 1989,
  window_label = "1985-1989"
)

baseline_90_94 <- make_baseline_fixed_window(
  df = analysis_data_europe,
  start_year = 1990,
  end_year = 1994,
  window_label = "1990-1994"
)

```

```{r, include = FALSE}
# adding region and first tax year to 5 year average dataset
baseline_85_89 <- baseline_85_89 %>%
  left_join(region_map, by = c("Entity", "Code")) %>%
  left_join(tax_years,   by = c("Entity", "Code"))

baseline_90_94 <- baseline_90_94 %>%
  left_join(region_map, by = c("Entity", "Code")) %>%
  left_join(tax_years,   by = c("Entity", "Code"))
```


```{r, include = FALSE}
# creating pre-tax averages for each variable in the 5 years before adopting the tax
pretax_levels <- analysis_data_europe %>%
  left_join(tax_years, by = c("Entity","Code")) %>%
  filter(!is.na(first_tax_year),
         year >= first_tax_year - 5,
         year <= first_tax_year - 1) %>%
  group_by(Entity, Code) %>%
  summarise(
    pretax_trade   = safe_mean(tradepercent),
    pretax_exports = safe_mean(exports_gdp),
    pretax_imports = safe_mean(imports_gdp),
    pretax_co2     = safe_mean(co2_per_capita),

    pretax_fossil  = safe_mean(fossil_share),
    fossil_obs     = sum(!is.na(fossil_share)),

    trade_obs   = sum(!is.na(tradepercent)),
    exports_obs = sum(!is.na(exports_gdp)),
    imports_obs = sum(!is.na(imports_gdp)),
    co2_obs     = sum(!is.na(co2_per_capita)),
    .groups = "drop"
  )
```


```{r, include = FALSE}
# calculating the carbon emissions trend for countries in the 5 years before they start the tax
pretax_co2_trend <- analysis_data_europe %>%
  left_join(tax_years, by = c("Entity","Code")) %>%
  filter(
    !is.na(first_tax_year),
    year >= first_tax_year - 5,
    year <= first_tax_year - 1,
    !is.na(co2_per_capita)
  ) %>%
  group_by(Entity, Code) %>%
  summarise(
    pretax_co2_trend = if (n() >= 2)
      coef(lm(co2_per_capita ~ year))[["year"]]
    else NA_real_,
    .groups = "drop"
  )
```

```{r, include = FALSE}
# joining calculated emissions trend to pre-tax averages dataset
pretax_baseline <- pretax_levels %>%
  left_join(pretax_co2_trend,   by = c("Entity","Code"))
```



```{r, include = FALSE}
# joining pretax averages and trends to dataset
analysis_data_europe <- analysis_data_europe %>%
  left_join(pretax_baseline, by = c("Entity","Code"))
```



```{r, include = FALSE}
# choosing event windows
pre_end <- 2004
k_full  <- 10
k_zoom  <- 5
early_cutoff <- 1994
```


```{r, include = FALSE}
# joining first tax years to dataset
analysis_data_europe <- analysis_data_europe %>%
  select(-any_of("first_tax_year")) %>%
  left_join(tax_years, by = c("Entity", "Code"))
```

```{r, include = FALSE}
# limiting to years pre-2004, and calculating event time
# adding log of gdp to data
df_pre <- analysis_data_europe %>%
  filter(year <= pre_end) %>%
  mutate(
    g = first_tax_year,
    treated_by_2004 = !is.na(g) & g <= pre_end,
    early_adopter = treated_by_2004 & g <= early_cutoff,

    event_time = ifelse(treated_by_2004, year - g, NA_real_),

    log_gdp = log(GDP_per_capita)
  ) %>%
  filter(treated_by_2004 | is.na(g) | g > pre_end) %>%
  filter(!is.na(co2_per_capita), !is.na(log_gdp), !is.na(annual_temp_c))%>%
  select(-first_tax_year)
```




```{r, include = FALSE}
# keep all observations within event window and any rows where event time isn’t defined
df_pre_fullwin <- df_pre %>%
  filter(is.na(event_time) | (event_time >= -k_full & event_time <= k_full))
```

```{r, include = FALSE}
# first full window event study
es_pre04_full <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1) + log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_fullwin,
  cluster = ~Entity
)
```



```{r, include = FALSE}
# first event study graph
iplot(es_pre04_full,
      ref.line = 0,
      main = paste0("Pre-2004 Carbon Tax Event Study (±", k_full, ")"),
      xlab = "Years since adoption",
      ylab = "Effect on CO₂ per capita (Tons)")
```



```{r, include = FALSE}
# creating a dataset measuring the amount of countries that contribute for each relative time period
treated_cov_full <- df_pre %>%
  filter(treated_by_2004, !is.na(g)) %>%
  filter(event_time >= -k_full, event_time <= k_full) %>%
  group_by(event_time) %>%
  summarise(
    n_treated_countries = n_distinct(Code),
    .groups = "drop"
  )
```

```{r, include = FALSE}
# graphing the relative time contribution dataset
support1 <- ggplot(treated_cov_full, aes(x = event_time, y = n_treated_countries)) +
  geom_col(width = 0.85) +
  theme_minimal() +
  labs(
    title = "Treated Country Data Contribution by Event Time",
    x = "Years since adoption (event time)",
    y = "Number of treated countries"
  )
```



```{r, include = FALSE}
# creating a dataset to zoom in the window between -5 and +5 event times
df_pre_zoom <- df_pre %>%
  filter(is.na(event_time) | (event_time >= -k_zoom & event_time <= k_zoom))
```


```{r, include = FALSE}
# zoomed in event study
es_pre04_zoom <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1) + log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_zoom,
  cluster = ~Entity
)

iplot(es_pre04_zoom,
      ref.line = 0,
      main = paste0("Pre-2004 Carbon Tax Event Study (±", k_zoom, ")"),
      xlab = "Years since adoption",
      ylab = "Effect on CO₂ per capita (Tons)")
```



```{r, include = FALSE}
# creating a dataset for only pre-1995 adopters
df_pre_early <- df_pre %>%
  filter(!treated_by_2004 | early_adopter)
```


```{r, include = FALSE}
# zooming into between -5 and +5 for the pre-1995 adopters
df_pre_early_zoom <- df_pre_early %>%
  filter(is.na(event_time) | (event_time >= -k_zoom & event_time <= k_zoom))
```



```{r, include = FALSE}
# pre-1995 adopters event study, with event times between -5 and +5
es_pre04_early_zoom <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1) + log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_early_zoom,
  cluster = ~Entity
)


iplot(es_pre04_early_zoom,
      ref.line = 0,
      main = paste0("Early adopters (≤", early_cutoff, "), Event Study (±", k_zoom, ")"),
      xlab = "Years since adoption",
      ylab = "Effect on CO₂ per capita (Tons)")
```



```{r, include = FALSE}
# switching NaN values to NA values
baseline_85_89 <- baseline_85_89 %>% mutate(across(where(is.numeric), ~ ifelse(is.nan(.x), NA, .x)))
baseline_90_94 <- baseline_90_94 %>% mutate(across(where(is.numeric), ~ ifelse(is.nan(.x), NA, .x)))
```

```{r, include = FALSE}
# creating a propensity score dataset from the 1985-89 average dataset
ps_85_89 <- baseline_85_89 %>%
  mutate(
    treated = early_adopter_1990_2004,
    log_gdp = log(baseline_gdp),
    region  = factor(region)
  ) %>%
  filter(
    gdp_obs  >= 3,
    co2_obs  >= 3
  ) %>%
  filter(
    !is.na(treated),
    !is.na(log_gdp),
    !is.na(baseline_co2),
    !is.na(co2_trend_baseline),
    !is.na(baseline_temp),
    !is.na(region)
  )
```


```{r, include = FALSE}
# estimating propensity scores on the 1985-89 propensity score dataset
ps_mod_85_89 <- glm(
  treated ~ log_gdp + baseline_co2,
  data = ps_85_89,
  family = binomial()
)

summary(ps_mod_85_89)
```

```{r, include = FALSE}
# weighting propensity scores from 1985-89 averages
ps_85_89 <- ps_85_89 %>%
  mutate(
    pscore = predict(ps_mod_85_89, type = "response"),
    w_att  = ifelse(treated == 1, 1, pscore / (1 - pscore))
  )
```


```{r, include = FALSE}
# graphing propensity score overlap for 1985-89 propensity scores
ggplot(ps_85_89, aes(x = pscore, fill = factor(treated))) +
  geom_density(alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Propensity Score Overlap (Baseline 1985–89)",
    x = "Propensity score",
    fill = "Early adopter (1990–2004)"
  )
```



```{r, include = FALSE}
# Weight distribution for 1985-89 propensity scores
ggplot(ps_85_89, aes(x = w_att)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(
    title = "ATT Weight Distribution (Baseline 1985–89)",
    x = "ATT weight"
  )
```



```{r, include = FALSE}
# creating a propensity score dataset from the 19990-1994 averages dataset
ps_90_94 <- baseline_90_94 %>%
  mutate(
    treated = early_adopter_1990_2004,
    log_gdp = log(baseline_gdp),
    region  = factor(region)
  ) %>%
  filter(
    gdp_obs  >= 3,
    co2_obs  >= 3
  ) %>%
  filter(
    !is.na(treated),
    !is.na(log_gdp),
    !is.na(baseline_co2),
    !is.na(co2_trend_baseline),
    !is.na(baseline_temp),
    !is.na(region)
  )
```

```{r, include = FALSE}
# estimating propensity scores for 1990-94 propensity score dataset
ps_mod_90_94 <- glm(
  treated ~ log_gdp + baseline_co2,
  data = ps_90_94,
  family = binomial()
)

summary(ps_mod_90_94)
```



```{r, include = FALSE}
# weighting propensity scores from 1990-94 averages
ps_90_94 <- ps_90_94 %>%
  mutate(
    pscore = predict(ps_mod_90_94, type = "response"),
    w_att  = ifelse(treated == 1, 1, pscore / (1 - pscore))
  )
```



```{r, include = FALSE}
# graphing propensity score overlap for the 1990-94 dataset
ggplot(ps_90_94, aes(x = pscore, fill = factor(treated))) +
  geom_density(alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Propensity Score Overlap (Baseline 1990–94)",
    x = "Propensity score",
    fill = "Early adopter (1990–2004)"
  )

```


```{r, include = FALSE}
# graphing weight distribution for 1990-94 propensity scores
ggplot(ps_90_94, aes(x = w_att)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(
    title = "ATT Weight Distribution (Baseline 1990–94)",
    x = "ATT weight"
  )
```

```{r, include = FALSE}
# joining propensity weights with event study data
df_pre_all <- df_pre %>%
  left_join(ps_85_89 %>% select(Code, w_att) %>% rename(w_att_85 = w_att),
            by = "Code") %>%
  left_join(ps_90_94 %>% select(Code, w_att) %>% rename(w_att_90 = w_att),
            by = "Code") %>%
  mutate(
    w_att_85 = ifelse(treated_by_2004, 1, w_att_85),
    w_att_90 = ifelse(treated_by_2004, 1, w_att_90)
  )
```



```{r, include = FALSE}
# event study for pre-1994 adopters using 1985-90 propensity score weights
es_early_w85 <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1) +
    log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_all %>%
    filter(
      (!treated_by_2004 | early_adopter) &
      (is.na(event_time) | (event_time >= -k_zoom & event_time <= k_zoom)) &
      !is.na(w_att_85)
    ),
  weights = ~ w_att_85,
  cluster = ~ Entity
)

iplot(
  es_early_w85,
  ref.line = 0,
  main = "Early adopters ES (±5), PS-weighted (baseline 1985–89)",
  xlab = "Years since adoption",
  ylab = "Effect on CO₂ per capita (Tons)"
)
```





```{r, include = FALSE}
# event study for pre-1994 adopters using 19990-94 propensity score weights
es_early_w90 <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1) +
    log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_all %>%
    filter(
      (!treated_by_2004 | early_adopter) &
      (is.na(event_time) | (event_time >= -k_zoom & event_time <= k_zoom)) &
      !is.na(w_att_90)
    ),
  weights = ~ w_att_90,
  cluster = ~ Entity
)

iplot(
  es_early_w90,
  ref.line = 0,
  main = "Early adopters ES (±5), PS-weighted (baseline 1990–94)",
  xlab = "Years since adoption",
  ylab = "Effect on CO₂ per capita (Tons)"
)
```



```{r, include = FALSE}
# zooming in to between -5 and +5 for fossil fuel analysis
# while filtering out missing fossil fuel data
df_pre_zoom_fossil <- df_pre_zoom %>%
  filter(!is.na(fossil_share))
```


```{r, include = FALSE}
# fossil fuel event study
es_fossil_zoom <- feols(
  fossil_share ~ sunab(g, year, ref.p = -1) +
    log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_zoom_fossil,
  cluster = ~Entity
)

iplot(es_fossil_zoom,
      ref.line = 0,
      main = "Effect of Carbon Tax on Fossil Fuel Share (±5)",
      xlab = "Years since adoption",
      ylab = "Effect on fossil fuel share (Percentage Points")

```




```{r, include = FALSE}
# contribution at each relative time for fossil fuel event study
df_pre_fossil_zoom <- df_pre %>%
  filter(
    is.na(event_time) | (event_time >= -k_zoom & event_time <= k_zoom)
  ) %>%
  filter(!is.na(fossil_share))

treated_fossil_cov <- df_pre_fossil_zoom %>%
  filter(!is.na(g)) %>%
  group_by(event_time) %>%
  summarise(
    n_countries = n_distinct(Code),
    .groups = "drop"
  )
```


```{r, include = FALSE}
# graphing the contributions at each relative event time in the fossil fuel event study
supportfossil <-ggplot(treated_fossil_cov, aes(x = event_time, y = n_countries)) +
  geom_col(width = 0.85) +
  theme_minimal() +
  labs(
    title = "Fossil Fuel Share - Treated cohort coverage by event time",
    subtitle = paste0("Number of treated countries contributing at each lead/lag (±", k_zoom, ")"),
    x = "Years since adoption (event time)",
    y = "Number of treated countries"
  )
```




```{r, include = FALSE}
# Zoom window (same as CO2 main result) for trade openness
df_pre_trade_zoom <- df_pre %>%
  filter(
    is.na(event_time) | (event_time >= -k_zoom & event_time <= k_zoom)
  ) %>%
  filter(!is.na(tradepercent))
```



```{r, include = FALSE}
# trade openness event study
es_trade_zoom <- feols(
  tradepercent ~ sunab(g, year, ref.p = -1) +
    log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_trade_zoom,
  cluster = ~Entity
)

iplot(
  es_trade_zoom,
  ref.line = 0,
  main = "Effect of Carbon Tax on Trade Openness (±5)",
  xlab = "Years since adoption",
  ylab = "Effect on trade openness (Percentage Points)"
)
```



```{r, include = FALSE}
# contributions at each relative time of the trade openness event study
treated_trade_cov <- df_pre_trade_zoom %>%
  filter(!is.na(g)) %>%
  group_by(event_time) %>%
  summarise(
    n_countries = n_distinct(Code),
    .groups = "drop"
  )
```

```{r, include = FALSE}
# graphing contributions at each relative time of the trade openness event study
supporttrade <- ggplot(treated_trade_cov, aes(x = event_time, y = n_countries)) +
  geom_col(width = 0.85) +
  theme_minimal() +
  labs(
    title = "Trade Openness - Treated cohort coverage by event time",
    subtitle = paste0("Number of treated countries contributing at each lead/lag (±", k_full, ")"),
    x = "Years since adoption (event time)",
    y = "Number of treated countries"
  )
```





```{r, include = FALSE}
# no controls event study using -5 to +5
es_pre04_zoom1 <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1)  |
    Entity + year,
  data = df_pre_zoom,
  cluster = ~Entity
)

iplot(es_pre04_zoom1,
      ref.line = 0,
      main = paste0("Pre-2004 Carbon Tax Event Study (no controls) (±", k_zoom, ")"),
      xlab = "Years since adoption",
      ylab = "Effect on CO₂ per capita (Tons)")
```



```{r, include = FALSE}
# pulling treated country codes to create a map
treated_codes <- df_pre_zoom %>%
  filter(!is.na(event_time)) %>%
  distinct(Code) %>%
  pull(Code)
```



```{r, include = FALSE}
# creating a map highlight countries within the analysis
countries_in_data <- analysis_data_europe %>%
  distinct(Entity) %>%
  pull(Entity)

countries_in_data <- recode(countries_in_data,
  "Czechia" = "Czech Republic",
  "United Kingdom" = "UK"
)

world <- map_data("world")

europe <- world %>%
  filter(long > -25, long < 45, lat > 34, lat < 72) %>%

  filter(!region %in% c("Morocco","Algeria","Tunisia","Libya","Egypt")) %>%
  mutate(in_dataset = region %in% countries_in_data)

map <- ggplot(europe, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = in_dataset), color = "gray60", size = 0.2) +
  coord_quickmap() +
  scale_fill_manual(values = c(`TRUE` = "#1f78b4", `FALSE` = "#e0e0e0"),
                    name = "In dataset?") +
  theme_void() +
  labs(title = "Countries in the Analysis")+
  theme(
  legend.position = "none",
  plot.title = element_text(hjust = 0.5, face = "bold")
)
```




## Abstract

This study examines whether the first carbon taxes in Europe reduced national carbon dioxide emissions. Using the Sun & Abraham method, I construct an event study using data from 29 European countries between 1985 and 2004. The results provide little evidence of an immediate impact of these taxes on carbon emissions on a macroeconomic level. Results also provide limited evidence of a modest reduction in fossil fuel share in the short run, and little evidence of a shift in trade openness in the same period. These finding suggest that, in the short run, early European carbon taxes may have been implemented at levels too low to generate a measurable decline in carbon emissions, despite influencing energy composition.


\newpage

## Introduction

As the world increasingly focuses on global emissions and their consequences, it is important to understand the policy tools that can be utilized to limit such emissions. An increasing number of countries are implementing these different political strategies, with varying levels of success. On a macroeconomic level, currently there are two mainstream mechanisms that countries use regarding carbon emissions: Taxes and Emissions Trading Systems (ETS). Much can be argued about whether they have been successful recently in causing decreasing emissions, but it is also important to understand the history of their effectiveness. Thus, we can discuss and correct our mistakes in the pursuit of a greener world. 
  
An Emissions Trading System allows a set amount of emission permits into a sector, or industry, and companies can freely trade permits as they like. The long-term goal then, would be to decrease the amount of permits available in each industry. Carbon taxes are more straightforward: they are an extra fee placed on emissions per ton. The first countries to implement taxes were in Europe in the early 1990s (Metcalf & Stock, 2023). This paper takes a closer look at those carbon taxes, as the EU ETS did not start until 2005, allowing for ample pre-ETS data on taxes, and hopefully setting the stage for a clearer isolation of the tax effect. The first taxes were initially not as large as the more current ones are; thus, I expect this analysis to find evidence of a null or small relationship between early European carbon taxes and country level emissions (Vaibhav, 2025). These countries may have started their carbon taxes in a trial period, allowing room for adjustment, making it so that early effects may not have been as pronounced, if at all.

## Literature Review

There is a growing section of literature focused on the effects of both carbon taxes and Emissions Trading Systems, with mixed results. A recent meta-analysis by Ahmad, Li, and Wu (2023) examined findings from 81 empirical papers published between 2011 and 2022, concluding that while many studies report emission reductions, other find no significant effects. Influentially, Andersson (2019) found that Sweden’s carbon tax that started in 1991 caused an 11 percent decline in transport-sector carbon emissions. However, these sector level emissions effects may not fully translate nationally on a macroeconomic level. These mixed results highlight the need for further studies into the history and effectiveness of carbon taxes and Emissions Trading Systems.

In policy analysis, it has become common to use the event study framework to evaluate policy adoptions. Importantly, in 2021, Sun and Abraham developed a new version of the event study estimator, dubbed the “Sun & Abraham” method. It allows for treated units to start treatment at different times, creating a more powerful analysis of a true treatment effect on units with heterogenous treatment times. This mirrors how countries in Europe began their tax policies. Using this new method, this paper looks back on European countries that began their taxes pre-2005, with a goal of adding new statistical techniques to the ongoing research on the macroeconomic effect of policies on carbon emissions.

## Data

Because Europe has the longest time span of carbon tax data, and most countries within using a tax, this study focuses there for the strongest regional comparisons. Data in this study come from several different sources, including World Bank, Global Carbon Budget, and Our World in Data. The final dataset spans 29 European countries across 20 years, from 1985 to 2004. It is in an unbalanced panel data form, with country-year observations. In total, there are 1,160 observations of 27 variables, including macroeconomic indicators, emissions data, and constructed variables used for identification. Table 1 includes the first 5 rows of the data, including the main important variables used for final identification.
  
```{r table1,echo=FALSE, warning = FALSE, results= 'asis'}
table1 <- analysis_data_europe %>%
  select(
    Country = Entity,
    Year = year,
    Temp = annual_temp_c,
    `GDP pc` = GDP_per_capita,
    Trade = tradepercent,
    `CO2 pc` = co2_per_capita,
    `Fossil Share` = fossil_share,
    Tax = carbon_tax
  )

preview <- head(table1, 5)

kable(preview, format = 'latex',
      caption = "Selected variables from the first five observations. Temperature is measured in degrees Celsius, GDP per capita in USD, Trade Openness as a percentage of GDP, CO2 Emissions in metric tons per capita, and Fossil Fuel Share as a percentage of total energy consumption. The tax variable indicates whether a country has a tax in a specific year, taking the value 1 if it does, and the value 0 if it does not. Missing values (NA) reflect unavailable data.") %>%
  kable_styling(latex_options = c("hold_position"))
```

\FloatBarrier

The main response variable used in this study is carbon dioxide emissions, measured in metric tons. The key treatment variable is a binary treatment indicator, which is equal to 1 for a country $i$, in year $t$ that does have a carbon tax. For a country-year observation that does not have a tax, the value of the variable would be equal to 0. Then, for some country $i$, the indicator variable would take on a value of 1 beginning in the first year it implements a carbon tax. Control variables include GDP per capita and annual average temperature, measured in degrees Celsius. These are important to include, as wealthier countries may be more likely to start a tax, while warmer countries may use their energy in different ways than colder ones, affecting their carbon emissions. 

The following variables were included in the dataset to check in later analyses, after the main specifificatian was evaluated. Imports and exports, each as a percentage of GDP are included, as well as trade openness (defined as the sum of exports plus imports as a share of GDP). These were included because the countries with high trade levels may respond differently to carbon taxes than low trading countries. Lastly, fossil fuel share is included; where it is defined as the percentage of the total energy consumption made up by fossil fuel usage. Countries might change how they use their fossil fuels after a tax, and that should be tracked. Implausible values in the fossil fuel share variable after 2014 were treated as missing. Table 2 below shows the summary statistics of the important chosen variables. It should be noted that large values of trade openness reflect small, trade-intensive countries, where exports and imports can exceed GDP.
  
  
```{r table2, echo=FALSE, warning=FALSE, results='asis'}
summarydata <- analysis_data_europe %>%
  dplyr::select(
    `Annual Temp (C)` = annual_temp_c,
    `GDP Per Capita (USD)` = GDP_per_capita,
    `Trade Openness (% of GDP)` = tradepercent,
    `CO2 Per Capita (metric tons)` = co2_per_capita,
    `Carbon Tax (=1 if present)` = carbon_tax,
    `Fossil Fuel Share (%)` = fossil_share
  ) %>%
  as.data.frame()

stargazer(
  summarydata,
  type = "latex",
  title = "Summary Statistics",
  summary.stat = c("mean", "sd", "min", "p25", "median", "p75", "max"),
  digits = 1,
  digits.extra = 0,
  header = FALSE
)
```


  
Of the 29 countries, 25 are current European Union countries. The 4 others are made up of Great Britain, Iceland, Switzerland, and Norway. These countries were chosen because being a part of the EU signified similar political and economic settings, sizes, and other unidentifiable similarities that are important for comparison. The 4 others were specifically included for important reasons: Great Britain, being a large, powerful western European country was a part of the EU until 2020 – strong evidence that it was and still is comparable politically and economically to the other EU countries. Norway, like its EU neighbors Sweden and Finland, was one of the first countries to implement a carbon tax in the world. Switzerland and Iceland both also belong due to their EU-resembling wealth levels and policy structures, and they are located within the western European scope of the other countries. This means that several eastern European countries, certain southern ones, and several small city-states are not included in the data. However, their political and economic differences to the EU countries make it such that excluding them leads to stronger identification. Figure 1 shows a map of the countries that are included in the analysis in blue.
  

```{r figure1,fig.align="center", echo=FALSE, warning=FALSE, fig.cap="Countries included in the analysis. Highlighted countries represent the 29 European nations used in the event study sample (1985–2004)."}
map
```

\FloatBarrier

The countries included in the analysis represent much of Western and Northern Europe, along with select countries from Central and Eastern Europe. Although it is not fully comprehensive, it captures the first adopters of carbon taxes and provides sufficient comparisons for those adopters. The panel structure of the data allows for comparisons of carbon emissions before and after tax implementation, relative to the countries that did not adopt a tax. The next section outlines the empirical strategy used to estimate this effect.



## Methodology

A common method for analyzing policy implementations is the event study, which focuses on estimating the impact over time, often in cases of policy analysis. It works much like a difference-in-difference model, only calculating differences for each time unit relative to the event start, rather than estimating one single post treatment effect. Relative time $k$ = 0 represents the time unit in which the event starts. In our case, $k$ = 0 is the year country $i$ starts its carbon tax. The following years take on relative values $k$ = 1, 2, 3, and so on. Preceding years take on negative values, where the year before $k$ = 0 is $k$ = -1; 2 years before the event is $k$ = -2, and the pattern continues. Event studies designate a specific relative year $k$ as a baseline year for comparison, usually $k$ = -1. $k$ = -1 is the baseline in this analysis as well. For each relative year $k$, the event study measures how much the difference between the treatment and control groups changes relative to the baseline year difference. This results in a coefficient $\beta_k$ estimating the average change in treatment-control differences between year $k$ and year $k$ = -1. For example, in time $k$ = 2, 2 years after the tax starting, the event study calculates the relative difference in the gap between treatment and control groups in year $k$ = 2 and the baseline difference in $k$ = -1. This results in an estimate for $\beta_2$. These $\beta_k$ coefficients, with their confidence intervals, can be plotted on a graph over relative time periods, tracing out the dynamic effect of the event over time on the response variable. 

The main assumption of the event study model is the parallel trends assumption: absent treatment, the treated group would follow a similar trend as the control group. If evidence for this assumption is strong, it can be argued that any dynamic effects found after baseline time $k$ = -1 are due to the impact of the event. To check for parallel trends, it is important to examine the $\beta_k$ coefficients in the time periods pre-event, the estimates for $k$ < 1. If they are near 0, that shows the gap between treatment and control stays relatively similar to the baseline period in the other pre periods; that is good evidence for strong parallel trends. If the pre-event estimates show a pattern, it becomes problematic for any causal inference. However, in this research, the heterogeneous starting tax years would bias an event study with just 1 starting year.
  
This is where the Sun & Abraham method can be applied: it allows for treated units to have different starting times using relative time scaling (Sun & Abraham). Consider Denmark, which started its carbon tax in 1992, and Estonia, which started its tax in 2000. We can assign relative starting time $k$ = 0 to Denmark for 1992, and Estonia, $k$ = 0 for 2000. Then, if we wanted to examine time $k$ = 1 for both countries, we would be analyzing 1993 for Denmark, and 2001 for Estonia. To calculate a $\beta_1$ estimate for $k$ = 1, we would then calculate the change in Denmark's treatment-control gap in 1993 with its baseline year 1991, and similarly, Estonia in 2001 with its respective baseline year 1999. In this way, the Sun & Abraham method calculates these difference-in-difference estimates for each treated country in year $k$ = 1, and takes their weighted average for an estimate of $\beta_1$. It is possible to construct these types of event studies using the sunab function in R’s fixest package. Figure 2 below shows the number of countries that contribute for each relative time coefficient.
  
```{r figure2, fig.cap="This figure shows the number of treated countries contributing observations at each event time. Close to relative time 0, there are a higher number of treated countries that have available data. However, close to relative time $k$ = -10, only later adopters have data stretching that far back, while only the first adopters have data that reaches relative time $k$ = 10."}
support1 
```

\FloatBarrier

There are more countries with available relative time data between time periods $k$ = -5 and $k$ = 5. This is where I will focus the scope of the event study. Thus, the study analyzes the relative years from 5 years before the tax to 5 years after the tax. Including any further times would lead to unstable estimates, with only a few contributing later adopters for earlier relative times, and only the first adopters for later relative times. Equation \ref{eq:event_study} shows the Sun & Abraham fixed effects equation used in the final model.
  

\begin{equation}
\label{eq:event_study}
\begin{aligned}
\text{CO}_2{}_{it} =\;& 
\sum_{k=-5}^{-2} \beta_k \mathbf{1}\{t - g_i = k\}
+ \sum_{k=0}^{5} \beta_k \mathbf{1}\{t - g_i = k\} \\
&+ \gamma_1 \log(\text{GDPpc}_{it})
+ \gamma_2 \text{Temp}_{it} \\
&+ \alpha_i + \lambda_t + \varepsilon_{it}
\end{aligned}
\end{equation}


The equation models an estimate for the carbon emissions for a country $i$ in a specific year $t$, with standard errors clustered at the country level. The $\beta_k$ terms are the coefficients of interest, representing the difference-in-difference estimate for each relative year $k$. Notice that the lead coefficients before the event go up until $k$ = -2, because $k$ = -1 is used as the baseline for other estimates. They are used to test parallel trends. The lag coefficients after the event begins cover from $k$ = 0 to $k$ = 5, and they are used to estimate dynamic effects. $\mathbf{1}\{t - g_i = k\}$ is the event time indicator function. For a country in time $t$, given their first tax year $g_i$, this function calculates that country’s relative time $k$ and turns only the $\beta_k$ associated with that $k$ on, multiplying all other $\beta_k$ coefficients by 0. Each $\beta_k$ indicates the difference in the gap between the treatment and control groups in year $k$ and baseline year $k$ = -1. $\gamma_1$ and $\gamma_2$ control for the log of the country-year GDP per capita and average annual temperature. The $\alpha_i$ represents country fixed effects, and $\lambda_t$ represents the year fixed effects. Finally, $\epsilon_{it}$ is the error term for each country-year, with expectation $\mathbb{E}[\varepsilon_{it} \mid X_{it}, \alpha_i, \lambda_t] = 0$.

Event studies in the same form but measuring the impact of the tax on trade openness, and fossil fuel share were also carried out, and more information on them can be found in the Possible Alternative Outomes discussion section later. For robustness, similar event studies were carried out with and without controls, as well as limiting to only pre-1994 adopters to make sure results were not heavily influenced by late adopters. These tests are discussed more in the Robustness Checks section.
  
## Results

The final model result is shown below, as discussed in the methodology section using Equation \ref{eq:event_study}. It is a Sun & Abraham event study with fixed effects.


```{r figure3, fig.cap="The resulting event study graph, where each $\\beta_k$ corresponds with a plotted value for each year. The x-axis plots the relative years, and y-axis plots the estimated effect on carbon emissions per capita. Each estimate is also plotted with its respective confidence interval; the point at $k = -1$ is a placeholder for the baseline, and not an actual estimate."}
es_pre04_zoom <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1) + log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_zoom,
  cluster = ~Entity
)

mainresult <- iplot(es_pre04_zoom,
      ref.line = 0,
      ci_level = 0.95,
      main = paste0("Pre-2004 Carbon Tax Event Study (±", k_zoom, ")"),
      xlab = "Years since adoption",
      ylab = "Effect on CO₂ per capita (Tons)")

abline(v = 0, lty = 2, col = "gray40")
```

\FloatBarrier

```{r table3, echo=FALSE, warning=FALSE, results='asis'}
etable(
  es_pre04_zoom,
  tex = TRUE,
  title = "Main Event Study Results",
  label = "tab:pre04_es",
  fitstat = ~ n + r2 + wr2,
  dict = c(
    "log_gdp" = "Log GDP per Capita",
    "annual_temp_c" = "Annual Temperature",
    "year = -5" = "Event Time = -5",
    "year = -4" = "Event Time = -4",
    "year = -3" = "Event Time = -3",
    "year = -2" = "Event Time = -2",
    "year = 0"  = "Event Time = 0",
    "year = 1"  = "Event Time = 1",
    "year = 2"  = "Event Time = 2",
    "year = 3"  = "Event Time = 3",
    "year = 4"  = "Event Time = 4",
    "year = 5"  = "Event Time = 5"
  ),
  digits = 3
)
```


\FloatBarrier

Table \ref{tab:pre04_es} shows the exact estimates for each relative year coefficient $k$, along with the controls. Notice that there is no estimate for $\beta_k$ at $k$ = -1, where that treatment-control difference is used as the baseline.

The pre-treatment coefficients are encouraging for the parallel trends assumption; we hope that the black dots in Figure 3 for relative times $k$ < -1 are near 0, the flat line running across the middle of the graph. None of the pre-event coefficients are significant, and Figure 3 shows that all 4 have confidence intervals that include 0, crossing over that horizontal line. This can be seen in Table \ref{tab:pre04_es}, where each pre-event coefficient is not statistically significant. Thus, the pre-event trend is mostly flat and near 0; all of this information is consistent with the parallel trends assumption.

However, in the post-event years, there are no patterns that suggest an immediate effect on carbon emissions. In Figure 3, all of the estimates for $k$ > -1 have confidence intervals that cross over 0, and only the estimate for time $k$ = 0 is significant at the 10 percent level. This is also reflected in Table \ref{tab:pre04_es}, where $\beta_k$ estimates for $k$ > -1 have large p-values. The estimate for $k$ = 0 is −0.1869, indicating that in the year of adoption, $\text{CO}_2$ per capita decreases by approximately 0.187 metric tons more in treated countries than in the control group, relative to the year prior to adoption. These post-event coefficients do not show any noticeable patterns and stay mostly flat, which can both be seen by their proximity to 0 in Figure 3, and large p-values in Table \ref{tab:pre04_es}. This shows that there is not strong evidence of an immediate effect of a carbon tax on national carbon emissions for European countries that adopted before 2004.

Elsewhere, the log of GDP shows a strong positive correlation to carbon emissions, which is statistically significant at the 1 percent level. Because GDP per capita is used in logarithmic form, its coefficient implies that a 1% increase in GDP per capita is associated with an approximate increase of 0.0527 metric tons of $\text{CO}_2$ per capita, holding other factors constant. This estimate should not be interpreted as causal. Annual temperature also shows a small negative correlation with carbon emissions, which is not statistically significant at any level.

## Robustness Checks

The first important robustness test I use is making sure the results hold without any control variables, to show their stability. This test uses Equation \ref{eq:event_study} without the log of GDP or annual average temperature as controls, and keeping time and country fixed effects. The output shown in Figure 4 shows that results hold, even without the included controls; parallel trends are strong, and there are no statistically significant or large discernible patterns in the time periods $k$ > -1. This suggests that the findings are not driven by the inclusion of control variables. The full estimated coefficients of this check can be found in Table 6 in the appendix. The model used is Equation \ref{eq:trade_event_study_nocontrols}, also found in the appendix.

```{r figure4, fig.align="center", echo=FALSE, fig.cap="Event study results without including log of GDP or annual average temperature. This result is very similar to Figure 3, which is evidence that the results are not fueled by the controls."}
es_pre04_zoom1 <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1)  |
    Entity + year,
  data = df_pre_zoom,
  cluster = ~Entity
)

iplot(es_pre04_zoom1,
      ref.line = 0,
      ci_level = 0.95,
      main = paste0("Pre-2004 Carbon Tax Event Study (no controls) (±", k_zoom, ")"),
      xlab = "Years since adoption",
      ylab = "Effect on CO₂ per capita (Tons)")

abline(v = 0, lty = 2, col = "gray40")
```

\FloatBarrier

As Figure 2 in the methodology section shows, there are a few countries with many pre-event relative time data available that can contribute to estimates of relative times $k$ < -5. Therefore, we must make sure that those countries are not heavily influencing the pre-event data and perhaps driving any results; this is important for the parallel trends assumption. For this test, I remove any adopters of the tax that adopted after 1994 and model results using Equation \ref{eq:event_study}, the same final model as the results section. If pre-event trends or results differ from the main result, that may be evidence that late adopters followed different trends than earlier ones, and parallel trends evidence would be less trustworthy. However, if the results hold, it would be evidence that including these later adopters does not skew the data, and it is reasonable to limit the scope of the event study from $k$ = -5 to $k$ = 5, the area of greatest contribution. Figure 5 below highlights that the overall results remain similar. Although the $\beta_k$ estimates change in position, they continue to have confidence intervals that cross over 0, the horizontal line. The estimate at $k$ = -2 is the only one that becomes significant. As the only pre-event estimate out of 4 to be significant, we can still be reasonably confident in the parallel trends assumption, which becomes dangerous if there is a visible systematic trend across pre-event estimates. The results give evidence for the parallel trends assumption, and post-event trends remain similar to the main result; results are robust to restricting the sample to earlier adopters. The full estimated model results are in Table 7 in the appendix.
	
```{r figure5, fig.align="center", echo=FALSE, fig.cap="Event study results after restricting the treated sample to adopters up to 1994. Overall, results remain similar: post-event estimates remain statistically insignificant with confidence intervals overlapping 0, while pre-event estimates do not display a systematic trend, despite the estimate at $k = -2$ becoming significant. This suggests that including later adopters after 1994 does not meaningfully bias results, supporting their includsion in the main model."}
es_pre04_early_zoom <- feols(
  co2_per_capita ~ sunab(g, year, ref.p = -1) + log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_early_zoom,
  cluster = ~Entity
)


iplot(es_pre04_early_zoom,
      ref.line = 0,
      ci_level = 0.95,
      main = paste0("Early adopters (≤", early_cutoff, "), Event Study (±", k_zoom, ")"),
      xlab = "Years since adoption",
      ylab = "Effect on CO₂ per capita (Tons)")

abline(v = 0, lty = 2, col = "gray40")
```

\FloatBarrier

## Possible Alternative Outcomes

Even though the main result suggests that the first carbon taxes did not influence carbon emissions in the immediate term, it remains unclear which possible channels could have been part of the reason why. In this section, 2 channels are examined: changes in fossil fuel dependence, and trade openness. Intuitively, if a carbon tax is placed on a country, that country might trade more, and outsource their carbon, lowering their own carbon emission levels. For fossil fuels, a country with a carbon tax may transition toward cleaner energy sources, which would be reflected in fossil fuel share. Given the null findings for emissions, examining these channels may provide insight into why they did not decline.

### Fossil Fuels

First, to analyze the effect of the carbon tax on fossil fuels, I use the same event study model as the main model, but set fossil fuel share as the response variable. Equation (\ref{eq:fossil_event_study}) shown below is used as the fossil fuel share model, where $\text{FossilShare}_{it}$ denotes fossil fuel energy consumption as a percentage of total energy use.

\begin{equation}
\label{eq:fossil_event_study}
\begin{aligned}
\text{FossilShare}_{it} =\;& 
\sum_{k=-5}^{-2} \beta_k \mathbf{1}\{t - g_i = k\}
+ \sum_{k=0}^{5} \beta_k \mathbf{1}\{t - g_i = k\} \\
&+ \gamma_1 \log(\text{GDPpc}_{it})
+ \gamma_2 \text{Temp}_{it} \\
&+ \alpha_i + \lambda_t + \varepsilon_{it}
\end{aligned}
\end{equation}


If a carbon tax is placed on a country, it might be expected that the country may see a decrease in the proportion of fossil fuel energy they consume out of their total energy consumption. Figure 6 displays the corresponding event study results. Full model results can be found in Table 4, in the appendix.
	
```{r figure6, fig.align="center", echo=FALSE, fig.cap="Event study on the impact of the carbon tax on fossil fuel share. Fossil fuel share is defined as the percentage of total energy consumption in a country that comes from fossil fuels. For relative years $k > 2$, estimates suggest a 2 percentage point decrease in fossil fuel share following the tax. However, the pre-event estimates for $k < -1$ are less encouraging; 2 are statistically significant, and relatively few countries contribute data in these earlier time periods."}
es_fossil_zoom <- feols(
  fossil_share ~ sunab(g, year, ref.p = -1) +
    log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_zoom_fossil,
  cluster = ~Entity
)

iplot(
  es_fossil_zoom,
  ref.line = 0,
  ci_level = 0.95,
  main = "Effect of Carbon Tax on Fossil Fuel Share (±5)",
  xlab = "Years since adoption",
  ylab = "Effect on fossil fuel share (Percentage Points)"
)

abline(v = 0, lty = 2, col = "gray40")
```

\FloatBarrier

These results show encouraging middling pre-event trends. 2 pre-event estimates are statistically significant, with confidence intervals that do not overlap 0, at $k$ = -2 and $k$ = -5. It should also be mentioned that there are a relatively limited amount of treated countries that contribute to each pre-event relative time period, for time periods $k$ < -1. Therefore, it is difficult to make strong conclusions from this specific study. More information about the amount of contributors in each time period can found in the appendix Figure 8. In the implementation year, $k$ = 0, fossil fuel share declines by approximately 1.4 percentage points more than the control group, relative to baseline year $k$ = -1. The next 2 years are not statistically different from 0, but the same drop shows up in the following 3 years, for $k$ > 2. These estimates approach almost -2 percentage points. This suggests that in the years $k$ = 3, 4, 5, the treated countries experience a decrease in fossil fuel share of approximately 2 percentage points greater than the control group, relative to the baseline year $k$ = -1. While avoiding strong conclusions, this is provides suggestive evidence of an effect of carbon taxes on fossil fuel share for these early adopters, potentially reflecting a shift toward cleaner energy sources. However, this effect may not have been large enough to influence total carbon emissions on the national level.

### Trade Openness

A similar event study is also used to track the effect of the tax on trade openness, where trade openness is defined as imports plus exports, as a percentage of total GDP. To outmaneuver a tax, producers in a country may to decide to import more, rather than emit carbon in their own production if it is more cost effective. Then after a carbon tax is implemented, an increase in trade openness may be observed, while domestic carbon emissions may stay constant or show signs of decreasing. Once again, the same equation as the main model is used, this time with trade openness as the response variable. For the trade openness model, Equation (\ref{eq:trade_event_study}) is shown below, and the full model output can be found in Table 5 in the appendix.

\begin{equation}
\label{eq:trade_event_study}
\begin{aligned}
\text{TradeOpen}_{it} =\;& 
\sum_{k=-5}^{-2} \beta_k \mathbf{1}\{t - g_i = k\}
+ \sum_{k=0}^{5} \beta_k \mathbf{1}\{t - g_i = k\} \\
&+ \gamma_1 \log(\text{GDPpc}_{it})
+ \gamma_2 \text{Temp}_{it} \\
&+ \alpha_i + \lambda_t + \varepsilon_{it}
\end{aligned}
\end{equation}

The log of GDP per capita is included at a control because GDP likely affects trade openness levels. For larger GDP countries, less trade is necessary due to domestic production capabilities. For smaller GDP countries, their economies may depend more on international trade, leading to high trade openness levels. Figure 7 displays the results.
	
```{r figure7, fig.align="center", echo=FALSE, fig.cap="Event study on the impact of the carbon tax on trade openness. Trade openness is defined as the sum of imports and exports as a percentage of GDP. Pre-event trends appear stable, with estimates having confidence intervals overlapping 0 in all relative time periods except for $k = -5$. Post-event coefficients remain statistically insignificant, with confidence intervals overlapping 0, indicating no evidence of an immediate effect of the tax on trade openness."}
es_trade_zoom <- feols(
  tradepercent ~ sunab(g, year, ref.p = -1) +
    log_gdp + annual_temp_c |
    Entity + year,
  data = df_pre_trade_zoom,
  cluster = ~Entity
)

iplot(
  es_trade_zoom,
  ref.line = 0,
  ci_level = 0.95,
  main = "Effect of Carbon Tax on Trade Openness (±5)",
  xlab = "Years since adoption",
  ylab = "Effect on trade openness (Percentage Points)"
)

abline(v = 0, lty = 2, col = "gray40")
```

\FloatBarrier

Pre-event trends appear reasonably flat, with the exception of an outlier at relative time $k$ = -5. Post-event trends remain close to 0 and statistically insignificant. This suggests that, for early European carbon tax adopters, the tax did not have an immediate effect on trade openness. This may be evidence for a possible explanation as for why national carbon levels did not decrease after the tax: producers still found it more cost effective to take on the cost of the tax and continue production, rather than outsourcing. While there is evidence of shifts in energy consumption, they may not have been large enough to fully influence carbon emissions on a macroeconomic level in the immediate term.

## Conclusion

This study examined the impact of the first European carbon taxes on national carbon emissions levels, using a Sun & Abraham event study with fixed effects. The results indicate no evidence of a statistically significant short-run reduction in national carbon emissions due to tax adoption. Similar finding emerge when examining trade openness, with no evidence of an effect in the 5 years following the tax implementation. While there is suggestive evidence that the taxes influenced energy composition, these changes may not have had the magnitude to directly influence national carbon emissions in the short run. 

This suggests that the earliest taxes that European countries adopted may have been too low to have a prompt impact on national carbon emissions. Sweden, whose carbon tax started in 1991, initially started with a tax of 27 US dollars per metric ton; currently, that tax has increased to nearly 140 US dollars per ton (Metcalf, 2019). Even after adjusting for inflation, the real value of the tax more than doubled over that time. This pattern is consistent with the interpretation that the first taxes may have been relatively modest and experimental to allow for adaptation, before adopters fully began to focus on decreasing emissions.

A primary limitation of this study is the use of a binary indicator for carbon tax adoption rather than incorporating variation in tax rates. The current specification does not account for differences in tax levels across countries or changes in tax rates over time. In addition, although the model design controls for country and year fixed effects, countries that adopt carbon taxes may differ systematically from non-adopters in ways that are not fully captured by the model, potentially influencing both policy adoption and emissions trends.

Future research could examine heterogeneity in tax rates, along with interactions with other climate policies, such as Emissions Trading Systems. The EU ETS phase IV began in 2021, and future research could also delve into the effects of these recent policy implementations. The continued analysis of climate policies is essential to understanding how to design them, in order to effectively reduce emissions and protect the global environment.

## References

Ahmad, M., Li, X. F., & Wu, Q. (2024). Carbon taxes and emission trading systems: Which one is more effective in reducing carbon emissions? A meta-analysis. Journal of Cleaner Production, 476, 143761.

Alper, A. E. (2017). Analysis of carbon tax on selected European countries: Does carbon tax reduce emissions. Applied Economics and Finance, 5(1), 29–36.

Andersen, M. S. (2016, December 1). An introductory note on carbon taxation in Europe: A Vermont briefing [Briefing paper]. Aarhus University; European Environment Agency (EEA).

Andersson, J. J. (2019). Carbon taxes and CO₂ emissions: Sweden as a case study. American Economic Journal: Economic Policy, 11(4), 1–30.

Im, H., & Kim, Y. H. (2022). Carbon tax or emission trading scheme: Which is better to reduce carbon emission? (Working Paper No. 4129081). Social Science Research Network. https://doi.org/10.2139/ssrn.4129081

Metcalf, G. E. (2019, May 3). Carbon taxes: What can we learn from international experience? EconoFact. https://econofact.org/carbon-taxes-what-can-we-learn-from-international-experience

Metcalf, G. E., & Stock, J. H. (2023). The macroeconomic impact of Europe’s carbon taxes: Online appendix [Online appendix]. https://ceepr.mit.edu/wp-content/uploads/2023/01/2023-002-Appendix.pdf

Sun, L., & Abraham, S. (2021). Estimating dynamic treatment effects in event studies with heterogeneous treatment effects. Journal of Econometrics, 225(2), 175–199.

Vaibhav, H. (2025, March 4). The history and legacy of carbon pricing around the world. Earth.Org. https://earth.org/the-history-and-legacy-of-carbon-pricing-around-the-world/


### Data Sources

Copernicus Climate Change Service. (2025). ERA5 monthly averaged data on single levels from 1940 to present (processed by Our World in Data) [Dataset]. https://cds.climate.copernicus.eu/

Dolphin, G., & Merkle, C. (2024). Covered by emissions trading system in at least one sector (processed by Our World in Data) [Dataset]. Resources for the Future, World Carbon Pricing Database.

Dolphin, G., & Merkle, C. (2024). Covered by tax instrument in at least one sector (processed by Our World in Data) [Dataset]. Resources for the Future, World Carbon Pricing Database.

Global Carbon Project. (2025). Global Carbon Budget (Version 15) (processed by Our World in Data) [Dataset].

Our World in Data. (2024). Population (based on various sources) [Dataset]. https://ourworldindata.org/

World Bank Group. (2024a). World development indicators: GDP per capita (constant 2015 US$) [Dataset]. https://data.worldbank.org/

World Bank Group. (2024b). World development indicators: Fossil fuel energy consumption (% of total) [Dataset]. https://data.worldbank.org/

World Bank Group. (2024c). World development indicators: Trade (% of GDP) [Dataset]. https://data.worldbank.org/

\newpage 

## Appendix

### Fossil Fuel Event Study Figures

```{r figure8, fig.align="center", echo=FALSE, fig.cap="Graph showing how many countries contribute to estimates for each relative time period in the fossil fuel event study. Notice the decrease in contributors in relative times $k$ < 1."}
ggplot(treated_fossil_cov, aes(x = event_time, y = n_countries)) +
  geom_col(width = 0.85) +
  theme_minimal() +
  labs(
    title = "Treated Country Data Contribution by Event Time: \n Fossil Fuel Share",
    x = "Years since adoption (event time)",
    y = "Number of treated countries"
  )
```
\FloatBarrier


```{r table4, echo=FALSE, warning=FALSE, results='asis'}
etable(
  es_fossil_zoom,
  tex = TRUE,
  title = "Fossil Fuel Event Study Results",
  label = "tab:fossil_zoom",
  fitstat = ~ n + r2 + wr2,
  dict = c(
    "log_gdp" = "Log GDP per Capita",
    "annual_temp_c" = "Annual Temperature",
    "year = -5" = "Event Time = -5",
    "year = -4" = "Event Time = -4",
    "year = -3" = "Event Time = -3",
    "year = -2" = "Event Time = -2",
    "year = 0"  = "Event Time = 0",
    "year = 1"  = "Event Time = 1",
    "year = 2"  = "Event Time = 2",
    "year = 3"  = "Event Time = 3",
    "year = 4"  = "Event Time = 4",
    "year = 5"  = "Event Time = 5"
  ),
  digits = 3
)
```

\FloatBarrier

### Trade Openness Event Study Figures

```{r figure9, fig.align="center", echo=FALSE, fig.cap="Graph showing how many countries contribute to estimates for each relative time period in the trade openness event study. The number of contributors remains consistent throughout the time periods."}
ggplot(treated_trade_cov, aes(x = event_time, y = n_countries)) +
  geom_col(width = 0.85) +
  theme_minimal() +
  labs(
    title = "Treated Country Data Contribution by Event Time: \n Trade Openness",
    x = "Years since adoption (event time)",
    y = "Number of treated countries"
  )
```

\FloatBarrier

```{r table5, echo=FALSE, warning=FALSE, results='asis'}
etable(
  es_trade_zoom,
  tex = TRUE,
  title = "Trade Openness Event Study Results",
  label = "tab:trade_zoom",
  fitstat = ~ n + r2 + wr2,
  dict = c(
    "log_gdp" = "Log GDP per Capita",
    "annual_temp_c" = "Annual Temperature",
    "year = -5" = "Event Time = -5",
    "year = -4" = "Event Time = -4",
    "year = -3" = "Event Time = -3",
    "year = -2" = "Event Time = -2",
    "year = 0"  = "Event Time = 0",
    "year = 1"  = "Event Time = 1",
    "year = 2"  = "Event Time = 2",
    "year = 3"  = "Event Time = 3",
    "year = 4"  = "Event Time = 4",
    "year = 5"  = "Event Time = 5"
  ),
  digits = 3
)
```

\FloatBarrier

\newpage

### Robustness Checks

Equation \ref{eq:trade_event_study_nocontrols} below shows the equation used for the robustness check excluding log of GDP per capita and annual average temperature.

\begin{equation}
\label{eq:trade_event_study_nocontrols}
\begin{aligned}
\text{CO}_2{}_{it} =\;&  
\sum_{k=-5}^{-2} \beta_k \mathbf{1}\{t - g_i = k\}
+ \sum_{k=0}^{5} \beta_k \mathbf{1}\{t - g_i = k\} \\
&+ \alpha_i + \lambda_t + \varepsilon_{it}
\end{aligned}
\end{equation}

```{r table6, echo=FALSE, warning=FALSE, results='asis'}
etable(
  es_pre04_zoom1,
  tex = TRUE,
  title = "Main Event Study Results with No Controls",
  label = "tab:no_controls",
  fitstat = ~ n + r2 + wr2,
  dict = c(
    "log_gdp" = "Log GDP per Capita",
    "annual_temp_c" = "Annual Temperature",
    "year = -5" = "Event Time = -5",
    "year = -4" = "Event Time = -4",
    "year = -3" = "Event Time = -3",
    "year = -2" = "Event Time = -2",
    "year = 0"  = "Event Time = 0",
    "year = 1"  = "Event Time = 1",
    "year = 2"  = "Event Time = 2",
    "year = 3"  = "Event Time = 3",
    "year = 4"  = "Event Time = 4",
    "year = 5"  = "Event Time = 5"
  ),
  digits = 3
)
```

```{r table7, echo=FALSE, warning=FALSE, results='asis'}
etable(
  es_pre04_early_zoom,
  tex = TRUE,
  title = "Main Event Study Results with Adopters Up to 1994",
  label = "tab:earliest_adopters",
  fitstat = ~ n + r2 + wr2,
  dict = c(
    "log_gdp" = "Log GDP per Capita",
    "annual_temp_c" = "Annual Temperature",
    "year = -5" = "Event Time = -5",
    "year = -4" = "Event Time = -4",
    "year = -3" = "Event Time = -3",
    "year = -2" = "Event Time = -2",
    "year = 0"  = "Event Time = 0",
    "year = 1"  = "Event Time = 1",
    "year = 2"  = "Event Time = 2",
    "year = 3"  = "Event Time = 3",
    "year = 4"  = "Event Time = 4",
    "year = 5"  = "Event Time = 5"
  ),
  digits = 3
)
```

\newpage

### Code

All code for this project is available at: https://github.com/evantoth1/SeniorExperience

### AI Appendix

The majority of my interactions with AI during this project came in the form of code debugging. I used ChatGPT, where I would paste in code errors, and receive feedback from the AI. I would go through this feedback, testing different approaches, and tailoring the code to fit my vision for the project. These ranged from data wrangling errors to applying the Sun & Abraham method in R, and implementing math notations and references in the correct form during the final document preparation.

The other way that I used ChatGPT was for support in understanding the Sun & Abraham method. After going through primary sources, I pasted relevant passages about the method into ChatGPT, asking clarifying questions about my interpretations of the model, and for feedback on my ideas about how I might apply this technique to my own project.

Overall, this allowed me to develop a stronger understanding of the method and how I could apply it, which I independently conveyed without AI in both my presentation and paper. All passages, interpretations, conclusions, and presentation slides were written by me and reflect my own ideas.

